---
title:  "Using Python in R Studio"
author:  "Perine - DATA 205"
date:  "March 2026"
#output: html_document
  
    
jupyter:
  jupytext:
    cell_metadata_filter: -all
    formats: ipynb,Rmd
    main_language: python
    text_representation:
      extension: .Rmdam
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.16.7
---

This RMarkdown notebook is prepared to show DATA 205 students how to use both to Python syntax in a familiar IDE. It modifies the [ISLP](https://www.statlearning.com/resources-python) notebook [Ch02-statlearn-lab.Rmd](https://github.com/intro-stat-learning/ISLP_labs/blob/stable/Ch02-statlearn-lab.Rmd) with configuration information and references to R syntax for comparison. It also incorporates part of an example from [How to Use R in Google Colab (2026 Update)](https://www.codingthepast.com/2023/10/03/R-in-GoogleColab.html)

#Setup
To run this demo lab, you will need two things:

* An installation of `Python3`, which is the specific version of `Python`  used in the labs.
* An installation of the  `reticulate` package.

You can download and install  `Python3`   by downloading installers and following the instructions available at [Python downloads](https://www.python.org/downloads/)

The `reticulate`provides an R interface to Python and enables conversion between R and Python in the same document. Python code chunks can be embedded and executed directly in Rmarkdown or Quarto documents. [Watch this video](https://www.youtube.com/watch?v=6541wEO_04o) for a general introduction on using `reticulate` to run Python in RStudio.

When it is installed, `reticulate` will look for Python installation. In RStudio, you can explicitly configure which Python interpreter to use by navigating to **Tools > Global Options > Python**

Once you have installed Python3 and configured the global options in RStudio, you can continue with the installation of the `reticulate` package:

```{r, setup}
#options(repos = c(CRAN = "https://cran.rstudio.com/"))
#install.packages("reticulate")
library(reticulate)
reticulate::py_config()
```

# Syntax Reminders:  R Vesus Python

## Basic Commands

In this demo, we will introduce some simple `Python` commands. 
 For more resources about `Python` in general, readers may want to consult the tutorial at [docs.python.org/3/tutorial/](https://docs.python.org/3/tutorial/). 
 

###Python is a calculator just like R

Addition and subtraction in R:
```{r}
3 + 5 - 1
```

Addition and subtraction in Python:

```{python}
3 + 5 - 1
```

Multiplication in R:
```{r}
2 * 402
```
Multiplication in Python:
```{python}
2 * 402
```
Multiplication and division in R:

```{r}
(139/5) * 2
```
Multiplication and division in Python:

```{python}
(139/5) * 2
```

In Python, a variable or feature is referred to as an object.  R is also an object-oriented language, but we tend to refer to variables as such, because of the orientation to statistical analysis. 

To assign objects, we use `=` instead of `<-`

Assigning values to two objects (variables), then performing an operation and displaying the result:
```{r}
gdp <- 592993954831
pop <- 100000000
gdp_per_capita<- gdp/pop
gdp_per_capita
```
Same operation, done in Python:
```{python}
gdp = 592993954831
pop = 100000000
gdp_per_capita = gdp/pop
print(gdp_per_capita)
```
##Arithmetic Operations on numbers:

Operator	  Name	      Description
a + b	    Addition	    Sum of a and b
a - b	   Subtraction	  Difference of a and b
a * b	  Multiplication	Product of a and b
a / b	True division	    Quotient of a and b
a // b	Floor division	Quotient of a and b, removing fractional parts
a % b	Modulus	Integer   remainder after division of a by b
a ** b	Exponentiation	a raised to the power of b
-a	  Negation	        The negative of a

##Concatenate versus numerical operation (addition)

In `Python`, textual data is handled using
*strings*. For instance, `"hello"` and
`'hello'`
are strings. 
We can concatenate them using the addition `+` symbol.

```{python}
"hello" + " " + "world"
```

 A string is actually a type of *sequence*: this is a generic term for an ordered list. 
 The three most important types of sequences are lists, tuples, and strings.  
We introduce lists now. 


The following command instructs `Python` to join together
the numbers 3, 4, and 5, and to save them as a
*list* named `x`. When we
type `x`, it gives us back the list.

```{python}
x = [3, 4, 5]
x
```

Note that we used the brackets
`[]` to construct this list. 

We will often want to add two sets of numbers together. It is reasonable to try the following code,
though it will not produce the desired results.

```{python}
y = [4, 9, 7]
x + y

```

The result may appear slightly counterintuitive: why did `Python` not add the entries of the lists element-by-element? 

 In `Python`, lists hold *arbitrary* objects, and  are added using  *concatenation*. 
 In fact, concatenation is the behavior that we saw earlier when we entered `"hello" + " " + "world"`. 
 
This example reflects the fact that 
 `Python` is a general-purpose programming language. Much of `Python`'s  data-specific
functionality comes from other packages, notably `numpy`
and `pandas`. 

Later in this lab, we will introduce the  `numpy` package. 
See [docs.scipy.org/doc/numpy/user/quickstart.html](https://docs.scipy.org/doc/numpy/user/quickstart.html) for more information about `numpy`.


##Data Types

Python has similar basic data types to R, but they have different names. You can determine the data type by cuing the command `type`:

```{python}
type("Hello World")
```

```{python}
type(42)
```

```{python}
type(54.3)
```

```{python}
type(False)
```

#Switching Between R and Python for Analysis

Here is a basic workflow in R to analyze the historical popularity of the baby name Bruno, using data from the `babynames` package in R.

```{r}
# Install the necessary package 
install.packages("babynames")

# Load the library
library(babynames)

# Inspect the structure of the dataset
str(babynames)

# Filter the dataset for a specific name
bruno_df <- babynames[babynames$name == "Bruno", ]

# Plot the historical popularity of the name
plot(x = bruno_df$year, 
    y = bruno_df$n,
    type = "l",
    main = "Historical Popularity of the Name Bruno in the US",
    xlab = "Year",
    ylab = "Number of Babies")
```

Now suppose that once you've extrated data from the `babynames` package, you prefer to do visualization and further analysis in Python.  Here's what a workflow might look like.

Initially in R:

```{r}
# Install the necessary package directly in the Colab cell
#install.packages("babynames")

# Load the library
#library(babynames)

# Inspect the structure of the dataset
str(babynames)

# Filter the dataset for a specific name
bruno_df <- babynames[babynames$name == "Bruno", ]

#Display the first few rows to make sure you have the desired data
head(bruno_df)


```

Now you have an R data frame.  We need to pull this R object into a Python.

Insert Python chunks (using green "C" icon or `Ctrl + Alt + I).  

```{python}


#Import the needed Python libraries
import pandas as pd
import seaborn as sns
import matplotlib.pyplot as plt
```

1.  Access the R data frame directly and convert it to a Pandas Data Frame

```{python}
df = r.bruno_df
```

2. Set the aesthetic style
```{python}
sns.set_theme(style="whitegrid")
```

3. Create a line plot showing the proportion of babies named Bruno over time


```{python}
plt.figure(figsize=(10, 6))
plot = sns.lineplot(data=df, x='year', y='prop', hue='sex')
```

4. Add titles and labels

```{python}
plot.set_title("Historical Popularity of the Name 'Bruno'", fontsize=16)
plot.set_xlabel("Year", fontsize=12)
plot.set_ylabel("Proportion of Total Births", fontsize=12)
```

5.  Display the plot

```{python}
plt.show()
```



